### 2024-09-27 Manuscript/main_ecography_R1.tex
Update main_ecography_R1.tex

First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 elements in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
These elements can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. 
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
We refer to this as the ``non-stressed trait space''.

@@

We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):

@@

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$ (i.e. removing the $n$ pairs of points that are at least $D$ distance away from each other to remove the desired number of elements).
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).

@@

\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e. an estimated number of categories summarising the distribution, like in a histogram).
Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$. 
For bandwidths with a low density (low values of $n$) we used $m>0$ to increase their resampling probability, and the opposite for bandwidths with a high density (high values of $n$; using $m<0$ to decrease sampling property).
This effectively flattened the normal distribution of the data by some factor $p$ (here we used $p=3$).
This approximates our \textbf{competition mechanism}.
\end{itemize}

@@

We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2 and 8 dimensional traits (supplementary materials).
We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g. for two spaces with 200 elements and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
\hline
  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of elements.\\
  Equalising fitness & Size change (``size'') & When species with trait combinations that are more extreme (i.e. species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient. \\
  Facilitation & Density change (``density'') & When species with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor. In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
  Environmental filtering & Position change (``position'') & When an increase in species trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable).\\
  Competitive exclusion & Evenness change (``evenness'') & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 elements resulting a trait space of 2, 4, or 8 dimensions.
We then applied different stressors to the resulting trait space.
We represent the elements removed by the stressors in blue and the ones kept in the trait space in orange. The first column illustrate the distribution of the elements in one dimension with the overlap between the removed (blue) and kept (orange) elements in brown (note we did not measure any metric on 1D spaces). The second column illustrate the same distributions in 2 dimensions. We did not illustrate the distributions in 4 or 8 dimensions here.
We used the following stressors:
B) \textbf{null mechanism}: randomly removing half of the elements;
C) \textbf{equalising mechanism}: removing elements on the edge of the distribution;
D) \textbf{facilitation mechanism}: removing elements to reduce the distance between pairs of elements (increasing local density; decreasing evenness);
E) \textbf{filtering mechanism}: removing elements on one extreme of the distribution;
F) \textbf{competition mechanism}: proportionally removing elements on the centre of the distribution (``flattening the curve'').}}
\label{Fig:simulations}
\end{figure}
\bigskip

@@

Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many others could have been used (e.g. \citealt{guillerme2020shifting}).
Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
The simulated trait spaces in 2, 4 or 8 dimensions had the same variance on all dimensions (i.e. we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g. PCA or PCO), the dimensions have by definition a decreasing variance.

>>>>

First, we simulated a single random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
This trait was simulated either as a 2, 4 or 8 dimensional trait, in other words a Brownian Motion simulated in a 2, 4 or 8 dimensions.
These observations can represent any biological focal entities, e.g. tips in a phylogenetic tree, individuals, species, populations, OTUs, etc. %@@@ link to glossary
This resulted in a neutral null model of trait evolution with no effect of competition, extinction, selection or other processes (\textit{sensu} \citealt{bausman2018modeling}).
We refer to this as the ``non-stressed trait space''.

@@

We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 observations).
In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):

@@

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$ (i.e. removing the $n$ pairs of points that are at least $D$ distance away from each other to remove the desired number of observations).
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).

@@

\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e. an estimated number of categories summarising the distribution, like in a histogram).
Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of observations in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$. 
For bandwidths with a low density (low values of $n$) we used $m>0$ to increase their resampling probability, and the opposite for bandwidths with a high density (high values of $n$; using $m<0$ to decrease sampling property).
This effectively flattened the normal distribution of the data by some factor $p$ (here we used $p=3$).
This approximates our \textbf{competition mechanism}.
\end{itemize}

@@

We repeated each simulation pipeline (generate a trait space and apply the stressor) for uncorrelated 4 dimensional traits (Figure \ref{Fig:simulation_results}) and uncorrelated 2 and 8 dimensional traits (supplementary materials).
We limited our simulations to a relatively small number of traits due to the constraints of some of the metrics used (e.g., \texttt{TPD::TPDsMean} is only implemented for up to 4 dimensions; \citealt{carmona2019trait}) but also to avoid dealing with the curse of dimensionality \citep{bellman1957dynamic}.
This curse changes the properties of space occupancy in a non-linear way depending on the element's distribution and the number of dimensions.
For example, the volume of a trait space (or hypervolume when using more than 3 dimensions) typically tends to zero in a high number of dimensions.
However, the rate at which it approaches zero is not linear and depends on the distribution of each element on every dimension.
This makes it practically difficult to compare two randomly generated spaces with similar characteristics (e.g. for two spaces with 200 observations and 10 dimensions generated in the exact same way, one might have a hypervolume of nearly 0 and the other one of $10^5$).
Note also that in our simulations, all dimensions have the same properties (i.e. same variance and distribution).
This is often not the case in empirical cases (e.g. see our empirical example) where the dimensions have a decreased variance due to ordination techniques.

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.3\linewidth}|p{0.3\linewidth}|p{0.4\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description}\\
\hline
  Null mechanism & Random removal (``random'') & No overall change in species traits or community structure except for the reduction in the number of observations.\\
  Equalising fitness & Size change (``size'') & When species with trait combinations that are more extreme (i.e. species located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient. \\
  Facilitation & Density change (``density'') & When species with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor. In other words, this simulates the idea that some species are more likely to be resilient if they are present in a community of species with shared traits rather than the opposite.\\
  Environmental filtering & Position change (``position'') & When an increase in species trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable).\\
  Competitive exclusion & Evenness change (``evenness'') & When functionally similar species compete more intensively with one another than with functionally dissimilar species (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting})\\

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 observations resulting a trait space of 2, 4, or 8 dimensions.
We then applied different stressors to the resulting trait space.
We represent the observations removed by the stressors in blue and the ones kept in the trait space in orange. The first column illustrate the distribution of the observations in one dimension with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces). The second column illustrate the same distributions in 2 dimensions. We did not illustrate the distributions in 4 or 8 dimensions here.
We used the following stressors:
B) \textbf{null mechanism}: randomly removing a proportion of the observations;
C) \textbf{equalising mechanism}: removing observations on the edge of the distribution;
D) \textbf{facilitation mechanism}: removing observations to reduce the distance between pairs of observations (increasing local density; decreasing evenness);
E) \textbf{observations mechanism}: removing observations on one extreme of the distribution;
F) \textbf{competition mechanism}: proportionally removing observations on the centre of the distribution (``flattening the curve'').}}
\label{Fig:simulations}
\end{figure}
\bigskip

@@

Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many others could have been used (e.g. \citealt{guillerme2020shifting}).
Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of observations in trait spaces in a non-random manner.
Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
The simulated trait spaces in 2, 4 or 8 dimensions had the same variance on all dimensions (i.e. we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g. PCA or PCO), the dimensions have by definition a decreasing variance.
